---
title: "Data Simulation: Mi Abogado Synthetic Cohort"
format: html
execute:
  warning: false
  message: false
---

## Overview

This document generates a synthetic dataset to replicate the core findings of Cooper, Doyle, and Hojman (2023). The program, *Mi Abogado*, provides enhanced legal representation to children in foster care. The randomization was stratified by age group (under 12 vs. 12 and older), sex, and region.

We simulate a cohort of $N = 1871$ children, matching the baseline summary statistics (Table 1) and injecting a treatment effect on the outcome variable: Days Living with Family (Table 3).

## 1. Environment Setup

```{r setup}
library(tidyverse)
# Set seed for reproducibility
set.seed(0331) 
```

## 2. Simulating Baseline Characteristics

First, we generate the 1,871 children and populate their baseline demographic and case characteristics based on the control group means from the paper's balance table.

```{r simulate-covariates}
n_obs <- 1871

synthetic_data <- tibble(
  child_id = 1:n_obs,
  
  # Demographics
  female = rbinom(n_obs, 1, 0.57), # 57% female
  age_at_rand = rnorm(n_obs, mean = 13.7, sd = 3.3), # Mean age 13.7
  
  # Stratification variables
  age_group = ifelse(age_at_rand < 12, "Under 12", "12 and Older"),
  region = sample(
    c("Metropolitan", "Valparaiso", "Biobio", "Maule"), 
    size = n_obs, 
    replace = TRUE, 
    prob = c(0.4, 0.2, 0.2, 0.2) # Approximated region distribution
  ),
  
  # Case Characteristics
  school_delay = rpois(n_obs, lambda = 0.8), # Mean delay 0.8 years
  allegation_neglect = rbinom(n_obs, 1, 0.85), # 85% neglect
  allegation_physical = rbinom(n_obs, 1, 0.26), # 26% physical abuse
  allegation_sexual = rbinom(n_obs, 1, 0.17) # 17% sexual abuse
) %>%
  # Ensure age doesn't dip into impossible negative bounds
  mutate(age_at_rand = pmax(age_at_rand, 6)) # Trial was for ages 6-18
```

## 3. Stratified Randomization

The intervention was capacity-constrained, so treatment assignment was randomized within strata.

```{r assign-treatment}
# Randomly assign ~31% to treatment (581 out of 1871 slots)
# Grouping by our strata to mimic the stratified block randomization
synthetic_data <- synthetic_data %>%
  group_by(region, age_group, female) %>%
  mutate(
    treated = sample(c(0, 1), size = n(), replace = TRUE, prob = c(0.69, 0.31))
  ) %>%
  ungroup()
```

## 4. Expanding to a Panel and Simulating Outcomes

The research relies on longitudinal administrative data, so we need to expand our cross-sectional data into a panel spanning 8 quarters (2 years post-randomization).

We use a Gamma distribution log-link structure to simulate "Days Living with Family". The control group averages ~21.7 days post-randomization, and the treatment group sees an increase of ~5.6 days. 

```{r simulate-panel}
# Expand to 8 quarters
panel_data <- synthetic_data %>%
  crossing(quarter = 1:8) %>%
  mutate(
    post = ifelse(quarter >= 4, 1, 0), # Assume randomization happens at end of Q3
    
    # Define the Data Generating Process (DGP) using a Gamma GLM structure
    # Baseline log-days, plus a small bump for older kids/females, 
    # and the explicit treatment effect in the post period.
    linear_predictor = 2.5 + 
                       0.1 * female + 
                       0.05 * (age_at_rand - 13) - 
                       0.2 * allegation_sexual + 
                       0.25 * treated * post, # The Treatment Effect (ATE)
                       
    # Translate predictor to the expected value (mu) via log-link inverse
    mu_days = exp(linear_predictor),
    
    # Introduce Gamma noise (shape/rate parameterization)
    # Shape = 2 gives a nice right-skewed distribution of days
    days_with_family = rgamma(n(), shape = 2, rate = 2 / mu_days)
  ) %>%
  # Cap days at 90 (max days in a quarter)
  mutate(days_with_family = pmin(round(days_with_family), 90))

# Check the simulated effect
panel_data %>%
  group_by(treated, post) %>%
  summarize(mean_days = mean(days_with_family), .groups = 'drop')
```

## 5. Export for Analysis

```{r export}
# Save the synthetic panel data for the replication and ML scripts
dir.create("data", showWarnings = FALSE)
write_rds(panel_data, "data/synthetic_mi_abogado_panel.rds")
```