---
title: "Baseline Replication: Difference-in-Differences"
format: html
execute:
  warning: false
  message: false
---

## Overview

In this section, we replicate the core Intent-to-Treat (ITT) econometric models from the evaluation of the Mi Abogado program. The primary outcome of interest is the number of days a child spends living with their biological or adoptive family in a given quarter.

We estimate two models:
1. **A Pooled Difference-in-Differences (DiD) Model:** To capture the average treatment effect across the entire post-randomization period.
2. **An Event Study Model:** To map the dynamic treatment effects quarter-by-quarter and verify the parallel trends assumption in the pre-treatment periods.

## 1. Environment Setup

```{r setup}
library(tidyverse)
library(fixest)       
library(modelsummary) 
library(broom)        
library(plotly)       

# Load the synthetic panel data generated in Chapter 1
panel_data <- read_rds("data/synthetic_mi_abogado_panel.rds")
```

## 2. Pooled Difference-in-Differences

We estimate the pooled ITT model where the outcome Y is regressed on strata covariates X, a treatment indicator T, a post-period indicator Post, and their interaction (T * Post). 

The coefficient on the interaction term represents the average difference between the treatment and control groups in the post-period relative to the pre-period. Standard errors are clustered at the child level to account for repeated observations.

```{r pooled-did}
# Run the pooled DiD using feols()
did_model <- feols(
  days_with_family ~ treated * post + female + age_group + region,
  data = panel_data,
  cluster = ~child_id # Clustering at the level of randomization
)

# Display a clean table
modelsummary(
  list("Days Living with Family" = did_model),
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Within",
  coef_rename = c("treated" = "Assigned to Mi Abogado (T)",
                  "post" = "Post-Randomization",
                  "treated:post" = "Treatment x Post (ATE)")
)
```

**Interpretation:** The coefficient on the interaction term represents the Average Treatment Effect (ATE). In our synthetic cohort, assignment to the Mi Abogado program increases the time a child spends living with their family by approximately 3.74 days per quarter ($p < 0.001$), mirroring the highly significant positive trajectory found in the original paper.

## 3. Dynamic Event Study

To understand how this effect evolves over time and to check for pre-treatment parallel trends, we estimate an event study specification. We interact the treatment indicator with each calendar quarter, omitting Quarter 3 (the period immediately prior to randomization) as our reference group.

```{r event-study}
# Run the event study model
# i() is a special fixest function that creates interactions and handles factors
es_model <- feols(
  days_with_family ~ treated +
    i(quarter) +
    i(quarter, treated, ref = 3) +
    female +
    age_group +
    region,
  data = panel_data,
  cluster = ~child_id
)

# Extract the coefficients for plotting
es_results <- broom::tidy(es_model, conf.int = TRUE) %>%
  # Fix: Look for both words separately so it catches any naming format
  filter(grepl("quarter", term) & grepl("treated", term)) %>%
  # Clean up the term column to just get the numeric quarter
  mutate(
    quarter = as.numeric(str_extract(term, "\\d+")),
    is_post = ifelse(quarter >= 4, "Post-Treatment", "Pre-Treatment")
  ) %>%
  # Add the reference quarter (Q3) back in with 0s for the plot
  bind_rows(
    tibble(
      term = "reference",
      estimate = 0,
      std.error = 0,
      conf.low = 0,
      conf.high = 0,
      quarter = 3,
      is_post = "Pre-Treatment"
    )
  ) %>%
  arrange(quarter)
```

## 4. Interactive Event Study Visualization

We use ggplot2 and plotly to create an interactive event study graph.

```{r plot-event-study}
# Build the base ggplot
p <- ggplot(es_results, aes(x = quarter, y = estimate, color = is_post)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  geom_vline(xintercept = 3.5, linetype = "solid", color = "darkred", alpha = 0.8) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, linewidth = 1) +
  scale_color_manual(values = c("Pre-Treatment" = "gray50", "Post-Treatment" = "steelblue4")) +
  scale_x_continuous(breaks = 1:8, labels = paste0("Q", 1:8)) +
  labs(
    title = "Event Study: Effect of Mi Abogado on Days Living with Family",
    subtitle = "Quarter 3 serves as the reference period.",
    x = "Quarter Relative to Randomization",
    y = "Estimated Effect (Days)",
    color = "Period"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )

# Convert to an interactive plotly object
ggplotly(p, tooltip = c("x", "y", "ymin", "ymax")) %>%
  layout(legend = list(orientation = "h", x = 0.3, y = -0.2))
```

**Interpretation:** The interactive plot above demonstrates a credible event study. The pre-treatment coefficients (Q1 through Q3) hover tightly around zero, validating the parallel trends assumption. Following randomization (indicated by the red line), we see a sharp and sustained increase in the number of days treated children spend living with their families.