---
title: "Causal ML Extension: Heterogeneous Treatment Effects"
format: html
execute:
  warning: false
  message: false
---

## Overview

The baseline difference-in-differences model identified an average treatment effect (ATE) of roughly 5.6 additional days living with family per quarter for children in the *Mi Abogado* program. However, child welfare interventions rarely impact all cases equally. The authors explore this heterogeneity by splitting the sample into subgroups. 

In this extension, we apply a machine learning approach to estimate the **Conditional Average Treatment Effect (CATE)**. Specifically, we use a Causal Forest  to understand not just *if* the program works on average, but *for whom* it works best.

$$ \tau(x) = E[Y_i(1) - Y_i(0) | X_i = x] $$

## 1. Environment Setup & Data Prep

```{r setup}
library(tidyverse)
library(grf)          
library(ggplot2)

# Load the synthetic panel data
panel_data <- read_rds("data/synthetic_mi_abogado_panel.rds")

# To use grf, we need cross-sectional data. We will aggregate the outcome 
# (average days with family) for each child in the post-randomization period.
cross_section <- panel_data %>%
  filter(post == 1) %>%
  group_by(child_id, treated, female, age_at_rand, age_group, region, 
           school_delay, allegation_neglect, allegation_physical, allegation_sexual) %>%
  summarize(avg_days_post = mean(days_with_family), .groups = 'drop')
```

## 2. Training the Causal Forest

The `grf` package requires our covariates ($X$) to be a numeric matrix. We will convert our categorical variables (like region) into dummy variables, separate our treatment indicator ($W$), and isolate our outcome ($Y$).

```{r train-forest}
# 1. Covariate Matrix (X)
# We use model.matrix to handle dummy encoding for the region and age_group factors
X_df <- cross_section %>%
  select(-child_id, -treated, -avg_days_post)

X <- model.matrix(~ . - 1, data = X_df)

# 2. Treatment Vector (W)
W <- cross_section$treated

# 3. Outcome Vector (Y)
Y <- cross_section$avg_days_post

# Train the Causal Forest
# We set a seed within the function for reproducibility
c_forest <- causal_forest(
  X = X,
  Y = Y,
  W = W,
  num.trees = 2000,
  seed = 0331
)

# Print out the Average Treatment Effect estimated by the forest
ate_cf <- average_treatment_effect(c_forest, target.sample = "all")
paste("Causal Forest ATE:", round(ate_cf["estimate"], 2), 
      "days (SE:", round(ate_cf["std.err"], 2), ")")
```

## 3. Variable Importance

Which baseline characteristics are most responsible for driving the heterogeneity in treatment effects? We can extract the variable importance scores from the causal forest to find out.

```{r var-importance}
# Extract variable importance
var_imp <- variable_importance(c_forest)
var_names <- colnames(X)

# Create a clean dataframe for plotting
importance_df <- tibble(
  Variable = var_names,
  Importance = as.numeric(var_imp)
) %>%
  arrange(desc(Importance)) %>%
  # Clean up variable names for the plot
  mutate(Variable = str_replace_all(Variable, "allegation_", "Allegation: "),
         Variable = str_replace_all(Variable, "region", "Region: "),
         Variable = str_to_title(str_replace_all(Variable, "_", " ")))

# Plot the importance
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "gray30", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Causal Forest: Variable Importance",
    subtitle = "Which factors drive the heterogeneity in treatment effects?",
    x = "Baseline Characteristic",
    y = "Importance Score"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

**Interpretation:** The plot above reveals which baseline characteristics the Causal Forest splits on most frequently to maximize differences in treatment effects. Age at Randomization overwhelmingly dominates the model, indicating it is the single most critical factor in predicting a child's response to the enhanced legal aid. To a much lesser but still notable extent, whether the child resides in the Metropolitan region and whether their case involves a Sexual abuse allegation also serve as key drivers in differentiating highly successful interventions from average ones.

## 4. The Distribution of Treatment Effects

Finally, we predict the specific CATE for every child in our dataset and plot the distribution. This highlights the spread of the program's effectiveness across the population.

```{r cate-distribution}
# Predict the CATE for each child (out-of-bag estimates)
cate_predictions <- predict(c_forest)$predictions
cross_section$CATE <- cate_predictions

# Plot the distribution of the CATEs
ggplot(cross_section, aes(x = CATE)) +
  geom_histogram(fill = "coral3", color = "white", bins = 30, alpha = 0.8) +
  geom_vline(xintercept = ate_cf["estimate"], linetype = "dashed", color = "black", linewidth = 1) +
  annotate("text", x = ate_cf["estimate"] + 0.2, y = 150, 
           label = "Average Treatment Effect", angle = 90, vjust = 1.5) +
  labs(
    title = "Distribution of Conditional Average Treatment Effects (CATE)",
    subtitle = "Each observation represents an individual child's estimated treatment effect.",
    x = "Estimated Increase in Days Living with Family per Quarter",
    y = "Count of Children"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

**Conclusion:** While the baseline model found an average benefit of 3.74 days, this CATE distribution reveals a striking bimodal (two-peaked) spread. The estimated individual treatment effects range from roughly 1 to over 6 additional days with family per quarter, with distinct clusters forming around the 3-day and 5-day marks. This proves that the program's benefits are highly non-uniform. By cross-referencing this distribution with our variable importance metrics, welfare agencies can mathematically identify the specific demographic profiles driving that high-impact 5-day peak, enabling data-driven targeting of constrained legal resources.