---
title: "Cost-Benefit Analysis: The Financial Impact"
format: html
execute:
  warning: false
  message: false
---

## Overview

Statistical significance is only one component of program evaluation; policy adoption ultimately requires financial justification. The *Mi Abogado* program requires a heavier upfront investment in legal aid, as its lawyers manage strictly capped caseloads and are paid at a higher rate than standard public defenders. 

In this section, we replicate the cost-benefit analysis from the original trial (Table 12) to determine if the program's success in reducing time in state custody offsets its operational costs.

## 1. Environment Setup & Cost Parameters

We begin by establishing the daily costs associated with the child welfare and legal systems in Chile, converted to 2022 US Dollars.

* **Legal Aid Costs:** The *Mi Abogado* program costs $4.99 per child per day, whereas standard non-MA legal representation costs $2.73 per child per day.
* **Substitute Care Costs:** Public, directly-administered residences are highly expensive at $67.27 per child per day. Nonprofit collaborative residences cost $28.35 per child per day, and nonprofit family foster care costs $13.94 per day.

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)

# Define the daily costs (in USD)
cost_ma_lawyer <- 4.99
cost_std_lawyer <- 2.73
cost_pub_residence <- 67.27
cost_np_residence <- 28.35
cost_foster_care <- 13.94
```

## 2. Estimating the Net Financial Impact

Over a nearly two-year observation period (721 days), children assigned to the treatment group experienced significant shifts in their care trajectories. We construct a financial summary table to calculate the net savings to the *Servicio Nacional de Menores* (SENAME) per child.

```{r cost-benefit-table}
# Replicating the exact treatment differences observed in the trial
cba_data <- tibble(
  Category = c(
    "Days w/ MA Lawyer",
    "Days w/ Non-MA Lawyer",
    "Days in residence (public)",
    "Days in residence (nonprofit)",
    "Days in family foster care"
  ),
  `Treatment Mean` = c(340.63, 92.42, 126.37, 321.83, 13.06),
  `Control Mean` = c(234.89, 182.00, 131.98, 351.29, 7.66),
  Difference = c(105.74, -89.57, -5.61, -29.46, 5.41),
  `Daily Cost ($)` = c(
    cost_ma_lawyer,
    cost_std_lawyer,
    cost_pub_residence,
    cost_np_residence,
    cost_foster_care
  )
) %>%
  mutate(
    `Net Impact ($)` = Difference * `Daily Cost ($)`
  )

# Calculate total savings
total_savings <- sum(cba_data$`Net Impact ($)`)

# Add a total row for the final table output
cba_table <- cba_data %>%
  bind_rows(
    tibble(
      Category = "Total Net Impact per Child",
      `Treatment Mean` = NA,
      `Control Mean` = NA,
      Difference = NA,
      `Daily Cost ($)` = NA,
      `Net Impact ($)` = total_savings
    )
  )

# Render a HTML table
options(knitr.kable.NA = '')

cba_table %>%
  kable(digits = 2, format.args = list(big.mark = ",")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(
    nrow(cba_table),
    bold = TRUE,
    color = "white",
    background = "steelblue4"
  )
```

## 3. Return on Investment (ROI)

The financial calculations reveal a compelling argument for scaling the program. 

**1. The Cost of Intervention:** The treatment group spent approximately 106 more days with a *Mi Abogado* lawyer and 90 fewer days with a standard lawyer. This resulted in an increased legal aid investment of roughly **$283.11 per child** ($527.64 - $244.53).

**2. The Savings from Permanency:** Because the program successfully accelerated family reunification, treated children spent significantly fewer days in highly expensive state and nonprofit residences. This reduction generated **$1,212.57 in savings per child** across the residential network.

**Conclusion:**
The *Mi Abogado* program easily pays for itself. The state generated an average net savings of **$854.04 per child**. Put another way, for every $1 of additional legal aid investment required to fund the program, the child welfare system saved over $3.00. This creates a rare policy scenario where improving child well-being simultaneously reduces government expenditure.